---
title: "DAT402Proj2"
author: "William Reid"
date: "2026-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(caret)
library(ggplot2)
library(cluster)
library(factoextra)
library(rpart)
library(rpart.plot)
library(nnet)
library(randomForest)
library(class)
```

Introduction

This project uses machine learning to predict a Pokemon’s primary type based on its stats, height, weight, and base experience.

The goal is to compare multiple machine learning models and determine which model performs best at classifying Pokemon types.

Dataset Description
```{r}
data <- read.csv("pokemon_data.csv")

head(data)
str(data)
summary(data)
```

Data Cleaning
```{r}
data_clean <- data %>%
  separate(
    stats,
    into = c("hp", "attack", "defense", "special_attack", "special_defense", "speed"),
    sep = ", "
  ) %>%
  mutate(
    hp = as.numeric(gsub("hp=", "", hp)),
    attack = as.numeric(gsub("attack=", "", attack)),
    defense = as.numeric(gsub("defense=", "", defense)),
    special_attack = as.numeric(gsub("special-attack=", "", special_attack)),
    special_defense = as.numeric(gsub("special-defense=", "", special_defense)),
    speed = as.numeric(gsub("speed=", "", speed)),
    primary_type = sapply(strsplit(types, ", "), `[`, 1),
    primary_type = as.factor(primary_type)
  ) %>%
  na.omit()

str(data_clean)
summary(data_clean)
```

# Exploratory Data Analysis

Type Distribution
```{r}
ggplot(data_clean, aes(x = fct_infreq(primary_type))) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Distribution of Pokemon Primary Types",
    x = "Primary Type",
    y = "Count"
  )
```

Attack vs Speed
```{r}
## Attack vs Speed by Type (Custom colors)

ggplot(data_clean, aes(x = attack, y = speed, color = primary_type)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c(
    "fire" = "red",
    "water" = "blue",
    "grass" = "green",
    "electric" = "gold",
    "psychic" = "purple",
    "rock" = "brown",
    "ground" = "tan",
    "ice" = "lightblue",
    "dragon" = "darkblue",
    "dark" = "black",
    "fairy" = "pink",
    "fighting" = "orange",
    "ghost" = "violet",
    "bug" = "darkgreen",
    "steel" = "gray"
  )) +
  labs(
    title = "Attack vs Speed by Primary Type",
    x = "Attack",
    y = "Speed",
    color = "Primary Type"
  )
```

Stats by Primary Type
```{r}
data_long <- data_clean %>%
  select(primary_type, hp, attack, defense, special_attack, special_defense, speed) %>%
  pivot_longer(
    cols = c(hp, attack, defense, special_attack, special_defense, speed),
    names_to = "stat",
    values_to = "value"
  )
``` 
```{r}
avg_stats <- data_long %>%
  group_by(primary_type, stat) %>%
  summarise(mean_value = mean(value), .groups = "drop")

ggplot(avg_stats, aes(x = primary_type, y = mean_value, fill = stat)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Average Pokemon Stats by Type",
    x = "Primary Type",
    y = "Average Value"
  )
```

Unsupervised Learning (PCA)
```{r}
numeric_data <- data_clean %>%
  select(hp, attack, defense, special_attack, special_defense, speed)

scaled_data <- scale(numeric_data)

pca_model <- prcomp(scaled_data)

summary(pca_model)
```
```{r fig.width=10, fig.height=7}
fviz_pca_ind(
  pca_model,
  geom = "point",
  col.ind = data_clean$primary_type,
  title = "PCA Plot of Pokemon Stats"
)
```

Train/Test Split
```{r}
set.seed(123)

train_index <- createDataPartition(data_clean$primary_type, p = 0.7, list = FALSE)

train <- data_clean[train_index, ]
test <- data_clean[-train_index, ]
```

Model 1: Decision Tree
```{r}
tree_model <- rpart(primary_type ~ hp + attack + defense + special_attack + special_defense + speed + height + weight + base_experience,
                    data = train,
                    method = "class")

rpart.plot(tree_model)

tree_pred <- predict(tree_model, test, type = "class")

tree_cm <- confusionMatrix(tree_pred, test$primary_type)
tree_cm
```

Model 2: Random Forest
```{r}
set.seed(123)

rf_model <- randomForest(
  primary_type ~ hp + attack + defense + special_attack + special_defense + speed + height + weight + base_experience,
  data = train,
  ntree = 500,
  importance = TRUE
)

rf_pred <- predict(rf_model, test)

rf_cm <- confusionMatrix(rf_pred, test$primary_type)
rf_cm
```
```{r}
varImpPlot(rf_model)
```

Model 3: KNN
```{r}
train_x <- train %>%
  select(hp, attack, defense, special_attack, special_defense, speed, height, weight, base_experience)

test_x <- test %>%
  select(hp, attack, defense, special_attack, special_defense, speed, height, weight, base_experience)

train_x_scaled <- scale(train_x)

test_x_scaled <- scale(
  test_x,
  center = attr(train_x_scaled, "scaled:center"),
  scale = attr(train_x_scaled, "scaled:scale")
)

knn_pred <- knn(
  train = train_x_scaled,
  test = test_x_scaled,
  cl = train$primary_type,
  k = 5
)

knn_cm <- confusionMatrix(knn_pred, test$primary_type)
knn_cm
```

Model 4: Neural Network
```{r}
set.seed(123)

nn_model <- nnet(
  primary_type ~ hp + attack + defense + special_attack + special_defense + speed + height + weight + base_experience,
  data = train,
  size = 5,
  maxit = 500,
  trace = FALSE
)

nn_pred <- predict(nn_model, test, type = "class")
nn_pred <- as.factor(nn_pred)

nn_cm <- confusionMatrix(nn_pred, test$primary_type)
nn_cm
```

Results
```{r}
results <- data.frame(
  Model = c("Decision Tree", "Random Forest", "KNN", "Neural Network"),
  Accuracy = c(
    tree_cm$overall["Accuracy"],
    rf_cm$overall["Accuracy"],
    knn_cm$overall["Accuracy"],
    nn_cm$overall["Accuracy"]
  )
)

results
```

```{r}
ggplot(results, aes(x = reorder(Model, Accuracy), y = Accuracy)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Model Accuracy Comparison",
    x = "Model",
    y = "Accuracy"
  )
```

Conclusion

This project used machine learning to predict a Pokemon's primary type based on its stats and physical characteristics.

The models compared were a decision tree, random forest, KNN, and neural network. Based on the accuracy results, the best-performing model was random forest.

The results show that Pokemon stats provide some useful information for predicting type, but there are limitations. Many Pokemon types overlap in their stats, which makes classification difficult.

Future improvements could include using more features, tuning model parameters, trying different neural network architectures, or simplifying the prediction problem by grouping similar Pokemon types together.


